Evolution of ORM Systems: Formal Foundations

### 2012-12-30 doc/description.tex
text about transformations, format updates in app model

The model uses a set $Labels$ which contains all possible identifiers for the model. Each model component which has a label ($label \in Labels$) is unambiguously identified in the model.

$$
Application = < Class* > 
$$
$classes : Application \rightarrow Class*$ \\
$classes(<cs, is >) = cs, cs \in Class* \wedge is \in Instance* $ \\ \\
$class : Label \times Application \rightarrow Class  $ \\ 
$class(l, a) = c, c \in classes(a) \wedge name(c) = l, l \in Label, a = Application $ \\ \\
The structure of an application consists not only of classes, it is defined by five other concepts. Besides the structure definition, there is defined a set of functions, which allows to manipulate the application model easily.

@@

$$Class = < label, Property*, Association* >$$
$name : Class \rightarrow Label$ \\
$name(< lab, props, assocs  >) = lab$ \\ \\
$properties : Class \rightarrow Property*$ \\
$properties(< lab, props, assocs  >) = props $\\ \\
$primitiveProperties : Class \rightarrow Property*$ \\
$primitiveProperties(< lab, props, assocs  >) = props', \forall p \in props' : cardinality(p) \leq 1 $\\ \\
$collectionProperties : Class \rightarrow Property*$ \\
$collectionProperties(< lab, props, assocs  >) = props', \forall p \in props' : cardinality(p) > 1 $\\ \\
$associations : Class \rightarrow Association*$ \\
$associations(< lab, props, assocs  >) = assocs $ \\ \\
$associated :  Class \times Application \rightarrow Boolean $
%TODO: variable types

\paragraph{Property} represents a feature of  a class which is can be represented as a primitive type. The property can be mandatory, can have a default value.
$$
Property = < label, AppType, DefaultValue, Cardinality, Mandatory >
$$
$name : Property \rightarrow Label$ \\
$name(< lab, t, val, n  >) = lab$ \\ \\
$type : Property \times AppType$ \\
$type(< lab, t, val, n  >) = t$ \\ \\
$cardinality : Property \rightarrow NzNat$ \\
$cardinality(< lab, t, val, n  >) = n$ \\ \\
$mandatory : Property \rightarrow Boolean $
$mandatory(<lab, t, val, n, m >) = = m $ \\ \\
$owningClass : Property \times Application \rightarrow Class $ \\
$owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) $
%TODO: variable types

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be a starting class of an association, referenced class is consider to be an ending class of an association. The cardinalities defines the multiplicity of the association.
$$
Association = (label, classRef, startCardinality, endCardinality)
$$
$name : Association \rightarrow Label$ \\
$name(< lab, ref, n_1, n_2  >) = lab$\\ \\
$reference : Association \rightarrow Label$ \\
$reference(< lab, ref, n_1, n_2  >) = ref$\\ \\
$startCardinality : Association \rightarrow NzNat$ \\
$startCardinality(< lab, ref, n_1, n_2  >) = n_1$\\ \\
$endCardinality : Association \rightarrow NzNat$ \\
$endCardinality(< lab, ref, n_1, n_2  >) = n_2$ \\ \\
$owningClass : Association \times Application \rightarrow Class $ \\
$owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) $
%TODO: variable types

\paragraph{AppType} represents primitive types in the application. There are usually defined types such as String, Integer, Boolean etc. in contrast there is only one type in our model, because we focus on structural and data changes and type casting operations are not important for us. The only App-Type type is called APP-STRING.
$$
AppType = APPSTRING
$$
The types used in application model:
$$
cardinality = NzNat
$$
$$
referencesClass = label
$$
$$
startCardinality = NzNat
$$
$$
endCardinality = NzNat
$$

@@

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
$name : Table \rightarrow Label $ \\
$name(< lab, pk, cols, fks  >) = lab$ \\ \\
$primaryKey : Table \rightarrow PrimaryKey $ \\
$primaryKey(< lab, pk, cols, fks  >) = pk$ \\ \\
$columns : Table \rightarrow Column* $ \\
$columns(< lab, pk, cols, fks  >) = cols$ \\ \\
$foreignKeys : Table \rightarrow ForeignKey* $ \\
$foreignKeys(< lab, pk, cols, fks  >) = fks$

\paragraph{Column} columns define possible data values and types which can be part of a table record.
$$
Column = (label, DbType, DefaultValue, Constraint*)
$$
$name : Column \rightarrow Label $ \\
$name(< lab, t, val, cons  >) = lab $ \\ \\
$type : Column \rightarrow DbType $ \\
$type(< lab, t, val, cons  >) = t $ \\ \\
$constraints : Column \rightarrow Constraint* $ \\
$constraints(< lab, t, val, cons  >) = cons $ \\ \\
$owningTable : Column \times Database \rightarrow Table $ \\
$constraints(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) $

\paragraph{Foreign key}
$$
ForeignKey = (label, tableRef, Constraint*)
$$
$name : ForeignKey \rightarrow Label $ \\
$name(< lab, tRef, cons  >) = lab $ \\ \\
$reference : ForeignKey \rightarrow Label $ \\
$reference(< lab, tRef, cons  >) = tRef $ \\ \\
$reference : ForeignKey \rightarrow Constraint* $ \\
$reference(< lab, tRef, cons  >) = cons $ \\ \\
$owningTable : ForeignKey \times Database \rightarrow Table $ \\
$owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d) $

@@

\subsection{Mapping of Classes}
We assume the primary key column is created automatically by database, therefore the primary key is always created with name Id and od NzNat type and properties with cardinality 0 or 1 are mapped into columns. The associations are ignored int this mapping as it is one of the partial mapping function used in eORM.

@@

The evolution of an application is defined as a set of operations which change a structure of the application. The definition of an operation consists of two parts - a conditions of operation feasibility and operation impact on the application structure. If the conditions of feasibility are not met, the operation cannot change the application structure - the software cannot evolve. This is important in case we have an evolution consisting of several evolution operations. These operations has to be proceed in concrete order and all of them has to change the software otherwise the software can become inconsistent from user point of view (the structure or stored data will not correspond with the users' reality). To assure the right behavior of evolution operations and to propagate operations fail to the user we extend the definition of a software to make it a monad. The extension consists in defining a software state and operations $ret$ and $evolve$, which is used as a concatenation of operations as well.
$$state = Consistent s | Inconsistent s, s \in Softwares$$

$$ret : Software \rightarrow State $$
\begin{equation*}
\begin{gathered}
	ret(s) = \begin{cases}
 Consistent s, consistent(s)
 \\
 Inconsistent s, ! consistent(s)
 \end{cases}
\end{gathered}
\end{equation*}

Each evolution operation has a definition: $Software \rightarrow State$ and all evolution operations creates a set called $Operations$, then a concatenation of evolution operations is defined as:

$$evolve : State \times Operations \rightarrow State  $$
\begin{equation*}
\begin{gathered}
 evolve(t, o) = \begin{cases}
 t, \; t = Inconsistent \; s \\
 o(s), \; t = Consistent \; s 
\end{cases}\\ 
t \in States, o \in Operatons, s \in Softwares
\end{gathered}
\end{equation*}

We decide to use following transcription to cover all aspects  of operation definition:

\begin{equation*}
\begin{gathered}
operation(parameters, software) = \begin{cases}
  \inference{conditions \; of \; feasibility}{impact \; on \; software \; structure}\\ \\
  \inference{otherwise}{Inconsistent \; software}
 \end{cases}
\end{gathered}
\end{equation*}

@@

\paragraph{Change Cardinality}
$$\rho_{CcP} : Property \times NzNat \times Application \rightarrow Property $$
$$\rho_{cardA} : Association \times NzNat \times Application \rightarrow Association $$

>>>>

The model uses a set $Labels$ which contains all possible identifiers for the model. Each model component which has a label ($label \in Labels$) is unambiguously identified in the model.
\begin{gather*}
  Application = < Class* >
  \\ 
  app : Class* \rightarrow  Application \\ \\
  classes : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  class : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label, a = Application, cs \in Class* 
  \\ \\
  associated :  Label \times Application \rightarrow Boolean \\
  owningClass : Property \times Application \rightarrow Class  \\ owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) \\
  owningClass : Association \times Application \rightarrow Class  \\
owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c)
\end{gather*}
The structure of an application consists not only of classes, it is defined by five other concepts. Besides the structure definition, there is defined a set of functions, which allows to manipulate the application model easily.

@@

$$Class = <label, Property*, Association*>$$
\begin{equation*}
\begin{alignat}{2}
cl : Label \times Property* \times Association* \rightarrow Class &\quad&\text{constructor}\\ \\
% TODO association has to be only to self (all associated has to exists)
name : Class \rightarrow Label &\quad&\text{name of the class}\\
name(cl(l, props, assocs)) = l \\ \\
properties : Class \rightarrow Property* &\quad&\text{all properties}\\
properties(cl(l, props, assocs)) = props \\ \\
primitiveProperties : Class \rightarrow Property* &\quad&\text{all primitive properties}\\
primitiveProperties(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
collectionProperties : Class \rightarrow Property* &\quad&\text{all collections}\\
collectionProperties(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
associations : Class \rightarrow Association* &\quad&\text{all associations}\\
associations(cl(l, props, assocs)) = assocs
\end{alignat}
\end{equation*}
$$l \in Label, props \in Properties, assocs \in Associations$$

\paragraph{Property} represents a feature of  a class which is can be represented as a primitive type. The property can be mandatory, can have a default value.
\begin{gather*}
Property = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
prop : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
name : Property \rightarrow Label \\
name(prop(l, t, val, n, m)) = l \\ \\
type : Property \times AppType \\
type(prop(l, t, val, n, m)) = t \\ \\
cardinality : Property \rightarrow \mathbb{N^{*}} \\
cardinality(prop(l, t, val, n, m)) = n \\ \\
mandatory : Property \rightarrow Boolean \\
mandatory(prop(l, t, val, n, m)) = m  \\ \\
l \in Label, t \in AppType, val \in String*, n \in \mathbb{N^{*}}, m \in Boolean
\end{gather*}

%TODO: variable types

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be a starting class of an association, referenced class is consider to be an ending class of an association. The cardinalities defines the multiplicity of the association.
\begin{gather*}
Association = <label, classRef, startCardinality, endCardinality> \\ \\
assoc : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association
name : Association \rightarrow Label \\
name(assoc(l, ref, n_1, n_2)) = l\\ \\
reference : Association \rightarrow Label \\
reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
startCardinality : Association \rightarrow \mathbb{N} \\
startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
endCardinality : Association \rightarrow \mathbb{N} \\
endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
l, ref \in Label,  n_1, n_2 \in \mathbb{N}
\end{gather*}

%TODO: variable types

\paragraph{AppType} represents primitive types in the application. There are usually defined types such as String, Integer, Boolean etc. in contrast there is only one type in our model, because we focus on structural and data changes and type casting operations are not important for us. The only App-Type type is called APP-STRING.
$$
AppType = APPSTRING
$$

@@

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
$name : Table \rightarrow Label $ \\
$name(< lab, pk, cols, fks)) = lab$ \\ \\
$primaryKey : Table \rightarrow PrimaryKey $ \\
$primaryKey(< lab, pk, cols, fks)) = pk$ \\ \\
$columns : Table \rightarrow Column* $ \\
$columns(< lab, pk, cols, fks)) = cols$ \\ \\
$foreignKeys : Table \rightarrow ForeignKey* $ \\
$foreignKeys(< lab, pk, cols, fks)) = fks$

\paragraph{Column} columns define possible data values and types which can be part of a table record.
$$
Column = (label, DbType, DefaultValue, Constraint*)
$$
$name : Column \rightarrow Label $ \\
$name(< lab, t, val, cons)) = lab $ \\ \\
$type : Column \rightarrow DbType $ \\
$type(< lab, t, val, cons)) = t $ \\ \\
$constraints : Column \rightarrow Constraint* $ \\
$constraints(< lab, t, val, cons)) = cons $ \\ \\
$owningTable : Column \times Database \rightarrow Table $ \\
$constraints(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) $

\paragraph{Foreign key}
$$
ForeignKey = (label, tableRef, Constraint*)
$$
$name : ForeignKey \rightarrow Label $ \\
$name(< lab, tRef, cons)) = lab $ \\ \\
$reference : ForeignKey \rightarrow Label $ \\
$reference(< lab, tRef, cons)) = tRef $ \\ \\
$reference : ForeignKey \rightarrow Constraint* $ \\
$reference(< lab, tRef, cons)) = cons $ \\ \\
$owningTable : ForeignKey \times Database \rightarrow Table $ \\
$owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d) $

@@

\subsection{Mapping of Classes}
We assume the primary key column is created automatically by database, therefore the primary key is always created with name Id and of the $\mathbb{N^{*}}$ type and properties with cardinality 0 or 1 are mapped into columns. The associations are ignored int this mapping as it is one of the partial mapping function used in eORM.

@@

The evolution of a software is defined as a set of transformations which change a structure of the software. The definition of an operation consists of two parts - a conditions of operation feasibility and operation impact on the application structure. If the conditions of feasibility are not met, the operation cannot change the application structure - the software cannot evolve. This is important in case we have an evolution consisting of several evolution operations. These operations has to be proceed in concrete order and all of them has to change the software otherwise the software can become inconsistent from user point of view (the structure or stored data will not correspond with the users' reality). To assure the right behavior of evolution operations and to propagate operations fail to the user we extend the definition of a software to make it a monad. The extension consists in defining a software state and operations $ret$ and $\oplus$ (called evolve), which is used as a concatenation of operations as well.
\begin{gather*}
state = Consistent \; s \: | \: Inconsistent \; s, v \in Softwares 
\end{gather*}

\begin{equation*}
\begin{gathered}
ret : Software \rightarrow State \\ \\
	ret(s) = \begin{cases}
 Consistent \; s, consistent(s)
 \\
 Inconsistent\;  s, ! consistent(s)
 \end{cases}
\end{gathered}
\end{equation*}
Each evolution transformation has a definition: $Software \rightarrow State$ and all evolution transformations creates a set called $Transformations$, then a concatenation of evolution transformation is defined as:
\begin{equation*}
\begin{gathered}
 \oplus : State \times Transformations \rightarrow State  \\ \\
 u \oplus t = \begin{cases}
 u, \; u = Inconsistent \; s \\
 t(u), \; u = Consistent \; s 
\end{cases}\\ 
u \in States, t \in Transformations, s \in Softwares
\end{gathered}
\end{equation*}
Each transformation consist in changes of the application and the database. To describe these two aspects od software evolution we decide to define transformations for each layer separately. The separation of transformation contexts help us to define transformation specific for application or database evolution, which can ease both - understanding and defining the transformation. The transformations specific for a context are called operations and they creates two sets- first one for application evolution: $AppOps$  and second one for database evolution $DbOps$. Applying of a transformation is defined:
\begin{gather*}
u \oplus t = ret(<application(u) \oplus o_{app}, database(u) \oplus o_{db}>) \\
u \in Softwares, t \in Transformations, o_{app} \in AppOps, o_{db} \in DbOps
\end{gather*}
The function $\Phi$ provides a mapping between the $AppOps$ and $DbOps$ sets. The $\Phi$ function makes possible to use a description of an application evolution as a source for the evolution of the whole software.
\begin{gather*}
\Phi : AppOps \rightarrow DbOps
\end{gather*}

The definition of an operation consists of two parts - a conditions of operation feasibility and operation impact on the context. We decide to use following transcription to cover all aspects  of operation definition:

\begin{equation*}
\begin{gathered}
operation(parameters, context) = \begin{cases}
  \inference{conditions \; of \; feasibility}{impact \; on \; context}\\ \\
  \inference{otherwise}{Inconsistent \; context}
 \end{cases}
\end{gathered}
\end{equation*}

@@

\paragraph{Change Cardinality}
$$\rho_{CcP} : Property \times \mathbb{N^{*}} \times Application \rightarrow Property $$
$$\rho_{cardA} : Association \times \mathbb{N^{*}} \times Application \rightarrow Association $$
